---
title: "16S rRNA Amplicon Sequencing Analysis Report"
author: "Automated QIIME 2 + Nextflow Pipeline"
date: "`r format(Sys.Date(), '%d %B %Y')`"

params:
  group_column: "Health_State"

output:
  word_document:
    toc: true
    toc_depth: 3
    fig_caption: true
  html_document:
    toc: true
    toc_depth: 3
    self_contained: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "6in"
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(tibble)
library(knitr)

group_column <- params$group_column

results_dir <- "results"
figures_dir <- "figures"
tables_dir <- "tables"
metadata_file <- "metadata.tsv"

read_metadata_safe <- function(file) {
  if (!file.exists(file)) {
    return(NULL)
  }

  tryCatch(
    read.delim(
      file,
      header = TRUE,
      sep = "\t",
      check.names = FALSE,
      stringsAsFactors = FALSE
    ),
    error = function(e) {
      NULL
    }
  )
}

metadata <- read_metadata_safe(metadata_file)

sample_count <- if (!is.null(metadata)) {
  nrow(metadata)
} else {
  NA
}

group_count <- if (!is.null(metadata) &&
                    group_column %in% names(metadata)) {
  dplyr::n_distinct(metadata[[group_column]])
} else {
  NA
}

file_status <- function(files) {
  data.frame(
    File = basename(files),
    Available = file.exists(files),
    stringsAsFactors = FALSE
  )
}

safe_read_table <- function(file) {
  if (!file.exists(file)) {
    return(NULL)
  }

  tryCatch(
    read.csv(
      file,
      check.names = FALSE,
      stringsAsFactors = FALSE
    ),
    error = function(e) {
      NULL
    }
  )
}
```

# 1. Study Overview

This report summarizes the results generated by the automated 16S rRNA amplicon sequencing analysis pipeline.

The workflow uses Nextflow to organize the analysis and QIIME 2 for sequence processing, denoising, phylogenetic analysis, taxonomic classification, diversity analysis, and differential-abundance analysis.

The biological grouping used for downstream comparisons is:

**`r group_column`**

# 2. Dataset Summary

```{r dataset-summary}
summary_table <- data.frame(
  Metric = c(
    "Number of samples",
    "Number of groups",
    "Grouping variable"
  ),
  Value = c(
    ifelse(is.na(sample_count), "Not available", sample_count),
    ifelse(is.na(group_count), "Not available", group_count),
    group_column
  )
)

kable(summary_table)
```

# 3. Sample Metadata

The sample metadata used for the analysis are summarized below.

```{r metadata-table}
if (!is.null(metadata)) {

  kable(
    metadata,
    caption = "Sample metadata"
  )

} else {

  cat("Metadata file was not available.")
}
```

## Group Distribution

```{r group-distribution}
if (!is.null(metadata) &&
    group_column %in% names(metadata)) {

  group_summary <- metadata %>%
    count(.data[[group_column]], name = "Sample_Count")

  names(group_summary)[1] <- "Group"

  kable(
    group_summary,
    caption = "Number of samples in each biological group"
  )

} else {

  cat("The specified grouping variable was not available.")
}
```
# 4. DADA2 Denoising and ASV Generation

DADA2 was used to process the paired-end sequencing reads and generate
amplicon sequence variants (ASVs).

The denoising stage produces the feature table, representative sequences,
and denoising statistics.

```{r denoising-files}
denoising_files <- file_status(
  file.path(
    results_dir,
    "denoised",
    c(
      "table.qza",
      "rep-seqs.qza",
      "denoising-stats.qza",
      "denoising-stats.qzv"
    )
  )
)

kable(
  denoising_files,
  caption = "DADA2 denoising output files"
)
```

## Denoising Statistics

The DADA2 denoising statistics are stored in:

`results/denoised/denoising-stats.qzv`

This QIIME 2 visualization can be opened separately for detailed
inspection of read retention and filtering.

# 5. Phylogenetic Analysis

Representative ASV sequences were aligned and used to construct a
phylogenetic tree.

The phylogenetic tree is required for phylogeny-aware diversity measures
such as Faith's Phylogenetic Diversity and UniFrac.

```{r phylogeny-files}
phylogeny_files <- file_status(
  file.path(
    results_dir,
    "phylogeny",
    c(
      "aligned-rep-seqs.qza",
      "masked-aligned-rep-seqs.qza",
      "unrooted-tree.qza",
      "rooted-tree.qza"
    )
  )
)

kable(
  phylogeny_files,
  caption = "Phylogenetic analysis outputs"
)
```

# 6. Taxonomic Classification

Taxonomy was assigned to the representative ASV sequences using the
configured QIIME 2 classifier.

The main taxonomy outputs are the taxonomy artifact and its QIIME 2
visualization.

```{r taxonomy-files}
taxonomy_files <- file_status(
  file.path(
    results_dir,
    "taxonomy",
    c(
      "taxonomy.qza",
      "taxonomy.qzv",
      "taxa-bar-plots.qzv"
    )
  )
)

kable(
  taxonomy_files,
  caption = "Taxonomic classification outputs"
)
```

## Taxonomic Visualization

The taxa bar plot visualization provides an interactive overview of
taxonomic composition across samples.

The QIIME 2 visualization is:

`results/taxonomy/taxa-bar-plots.qzv`

# 7. Core Diversity Analysis

Core diversity analysis was performed using the QIIME 2 diversity workflow.

The analysis generated both alpha-diversity and beta-diversity metrics.

## Alpha Diversity Metrics

The generated alpha-diversity metrics include:

- Observed Features
- Shannon diversity
- Faith's Phylogenetic Diversity
- Evenness

```{r alpha-files}
alpha_files <- file_status(
  file.path(
    results_dir,
    "diversity",
    "core-metrics-results",
    c(
      "observed_features_vector.qza",
      "shannon_vector.qza",
      "faith_pd_vector.qza",
      "evenness_vector.qza"
    )
  )
)

kable(
  alpha_files,
  caption = "Alpha diversity outputs"
)
```

## Beta Diversity Metrics

The generated beta-diversity analyses include:

- Bray-Curtis
- Jaccard
- Weighted UniFrac
- Unweighted UniFrac

```{r beta-files}
beta_files <- file_status(
  file.path(
    results_dir,
    "diversity",
    "core-metrics-results",
    c(
      "bray_curtis_distance_matrix.qza",
      "bray_curtis_pcoa_results.qza",
      "jaccard_distance_matrix.qza",
      "jaccard_pcoa_results.qza",
      "weighted_unifrac_distance_matrix.qza",
      "weighted_unifrac_pcoa_results.qza",
      "unweighted_unifrac_distance_matrix.qza",
      "unweighted_unifrac_pcoa_results.qza"
    )
  )
)

kable(
  beta_files,
  caption = "Beta diversity outputs"
)
```
# 8. Beta Diversity PCoA Visualizations

Principal Coordinates Analysis (PCoA) was used to visualize differences
in microbial community composition between samples.

## 8.1 Bray-Curtis PCoA

Bray-Curtis PCoA represents differences in community composition based on
taxon abundance.

```{r bray-curtis-pcoa}
bray_plot <- file.path(
  figures_dir,
  "bray_curtis_pcoa.png"
)

if (file.exists(bray_plot)) {

  knitr::include_graphics(bray_plot)

} else {

  cat("Bray-Curtis PCoA figure has not been generated yet.")
}
```

## 8.2 Jaccard PCoA

Jaccard PCoA represents differences based primarily on the presence or
absence of features.

```{r jaccard-pcoa}
jaccard_plot <- file.path(
  figures_dir,
  "jaccard_pcoa.png"
)

if (file.exists(jaccard_plot)) {

  knitr::include_graphics(jaccard_plot)

} else {

  cat("Jaccard PCoA figure has not been generated yet.")
}
```

## 8.3 Weighted UniFrac PCoA

Weighted UniFrac incorporates both phylogenetic relationships and
relative abundance.

```{r weighted-unifrac-pcoa}
weighted_plot <- file.path(
  figures_dir,
  "weighted_unifrac_pcoa.png"
)

if (file.exists(weighted_plot)) {

  knitr::include_graphics(weighted_plot)

} else {

  cat("Weighted UniFrac PCoA figure has not been generated yet.")
}
```

## 8.4 Unweighted UniFrac PCoA

Unweighted UniFrac incorporates phylogenetic relationships while focusing
on presence and absence.

```{r unweighted-unifrac-pcoa}
unweighted_plot <- file.path(
  figures_dir,
  "unweighted_unifrac_pcoa.png"
)

if (file.exists(unweighted_plot)) {

  knitr::include_graphics(unweighted_plot)

} else {

  cat("Unweighted UniFrac PCoA figure has not been generated yet.")
}
```

# 9. Differential Abundance Analysis

Differential-abundance analysis was performed using ANCOM-BC.

The analysis was performed at both:

- ASV level
- Genus level

## 9.1 ASV-Level ANCOM-BC

```{r ancombc-asv-plot}
asv_plot <- file.path(
  figures_dir,
  "ancombc_asv_lfc.png"
)

if (file.exists(asv_plot)) {

  knitr::include_graphics(asv_plot)

} else {

  cat("ASV-level ANCOM-BC figure is not available.")
}
```

### ASV-Level Results

```{r ancombc-asv-results}
asv_results <- safe_read_table(
  file.path(
    tables_dir,
    "ancombc_asv_results.csv"
  )
)

if (!is.null(asv_results)) {

  kable(
    head(asv_results, 20),
    caption = "ASV-level ANCOM-BC results"
  )

} else {

  cat("ASV-level ANCOM-BC results table is not available.")
}
```

## 9.2 Genus-Level ANCOM-BC

```{r ancombc-genus-plot}
genus_plot <- file.path(
  figures_dir,
  "ancombc_genus_lfc.png"
)

if (file.exists(genus_plot)) {

  knitr::include_graphics(genus_plot)

} else {

  cat("Genus-level ANCOM-BC figure is not available.")
}
```

### Genus-Level Results

```{r ancombc-genus-results}
genus_results <- safe_read_table(
  file.path(
    tables_dir,
    "ancombc_genus_results.csv"
  )
)

if (!is.null(genus_results)) {

  kable(
    head(genus_results, 20),
    caption = "Genus-level ANCOM-BC results"
  )

} else {

  cat("Genus-level ANCOM-BC results table is not available.")
}
```
# 10. ANCOM-BC Summary

The ANCOM-BC analysis summary is provided below.

```{r ancombc-summary}
ancombc_summary <- safe_read_table(
  file.path(
    tables_dir,
    "ancombc_summary.csv"
  )
)

if (!is.null(ancombc_summary)) {

  kable(
    ancombc_summary,
    caption = "ANCOM-BC analysis summary"
  )

} else {

  cat("ANCOM-BC summary table is not available.")
}
```

# 11. Group Summary

The following table summarizes the number of samples represented in
each biological group.

```{r final-group-summary}
group_summary_file <- file.path(
  tables_dir,
  "group_summary.csv"
)

final_group_summary <- safe_read_table(group_summary_file)

if (!is.null(final_group_summary)) {

  kable(
    final_group_summary,
    caption = "Sample group summary"
  )

} else if (!is.null(metadata) &&
           group_column %in% names(metadata)) {

  final_group_summary <- metadata %>%
    count(.data[[group_column]], name = "Sample_Count")

  names(final_group_summary)[1] <- "Group"

  kable(
    final_group_summary,
    caption = "Sample group summary"
  )

} else {

  cat("Group summary is not available.")
}
```

# 12. Results Inventory

The pipeline results directory contains the QIIME 2 artifacts,
visualizations, and downstream analysis outputs generated during the
workflow.

```{r results-inventory}
inventory_file <- file.path(
  tables_dir,
  "results_inventory.csv"
)

inventory <- safe_read_table(inventory_file)

if (!is.null(inventory)) {

  kable(
    inventory,
    caption = "Generated results inventory"
  )

} else {

  all_files <- list.files(
    results_dir,
    recursive = TRUE,
    full.names = TRUE
  )

  inventory <- data.frame(
    File = all_files,
    stringsAsFactors = FALSE
  )

  kable(
    head(inventory, 100),
    caption = "Results inventory"
  )
}
```

# 13. Generated Figures

The following figures are generated by the R plotting component of the
pipeline.

```{r figure-inventory}
figure_files <- list.files(
  figures_dir,
  pattern = "\\.(png|jpg|jpeg)$",
  full.names = TRUE,
  recursive = TRUE
)

if (length(figure_files) > 0) {

  figure_table <- data.frame(
    Figure = basename(figure_files),
    Path = figure_files,
    stringsAsFactors = FALSE
  )

  kable(
    figure_table,
    caption = "Figures generated for the report"
  )

} else {

  cat("No report figures were generated.")
}
```

# 14. Generated Tables

```{r table-inventory}
table_files <- list.files(
  tables_dir,
  pattern = "\\.(csv|tsv)$",
  full.names = TRUE,
  recursive = TRUE
)

if (length(table_files) > 0) {

  table_table <- data.frame(
    Table = basename(table_files),
    Path = table_files,
    stringsAsFactors = FALSE
  )

  kable(
    table_table,
    caption = "Tables generated for the report"
  )

} else {

  cat("No report tables were generated.")
}
```

# 15. QIIME 2 Output Summary

The major stages of the analysis pipeline are summarized below.

| Analysis stage | Main outputs |
|---|---|
| Import | Demultiplexed sequence artifact |
| Denoising | Feature table, representative sequences, DADA2 statistics |
| Phylogeny | Aligned sequences and phylogenetic tree |
| Taxonomy | Taxonomic assignments and taxa visualization |
| Alpha diversity | Observed Features, Shannon, Faith's PD, Evenness |
| Beta diversity | Bray-Curtis, Jaccard, Weighted UniFrac, Unweighted UniFrac |
| Differential abundance | ANCOM-BC ASV and genus results |

QIIME 2 `.qza` files contain analysis artifacts, while `.qzv` files
contain interactive visualizations.

# 16. Conclusion

This report provides an automated summary of the 16S rRNA amplicon
sequencing analysis workflow.

The analysis includes sequence denoising, ASV generation, phylogenetic
reconstruction, taxonomic classification, alpha diversity, beta
diversity, and differential-abundance analysis.

The interpretation of biological group differences should consider
sample size, study design, sequencing depth, taxonomic assignment
confidence, and statistical results.

The original QIIME 2 `.qzv` files can be opened separately for detailed
interactive visualization and inspection.

# 17. Report Generation Information

This report was generated automatically from the results produced by
the Nextflow-based 16S rRNA amplicon analysis workflow.

The report contains both pipeline-generated tables and figures and
references to the original QIIME 2 artifacts.

```{r report-files}
report_files <- data.frame(
  Component = c(
    "Metadata",
    "Results directory",
    "Figures directory",
    "Tables directory"
  ),
  Path = c(
    metadata_file,
    results_dir,
    figures_dir,
    tables_dir
  ),
  Available = c(
    file.exists(metadata_file),
    dir.exists(results_dir),
    dir.exists(figures_dir),
    dir.exists(tables_dir)
  )
)

kable(
  report_files,
  caption = "Report input and output locations"
)
```
